# **Mineralogy and Clay Minerals**

::: callout-note
## Objectives

-   Draw the structure of clay minerals.
-   Understand why the structure of the clay determines the soils physical and chemical properties.
-   Describe differences between 1:1 and 2:1 clay minerals.
:::

::: callout-tip
## Key Words & Concepts

-   Kaolinite
-   Illite
-   1:1
-   Octahedral
-   Smectite
-   Expanding clay
-   Tetrahedral
-   Vermiculite
-   2:1
:::

## INVESTIGATION A: 3-D Digitized and Model Clay mineral structures

Using the iPads provided, choose the iBook named SOIL 2125 Mineral Structures- Lab 2. Start on the first page, and flip through the iBook. Be sure to read the captions. Turn pages by swiping right or left on the top or bottom. The figures are fully-rotatable 3-D diagrams. You can use your finger on the screen to manipulate them. If you need assistance, ask the TA. **Please navigate with the bookmark or flip back to the very first page (Interactive 1.1) for the next person when you are done.**

Answer the following questions from your observation and reading of the models in the iBook: 
1. What other elements are present in the mineral structure of Feldspars that are not present in Quartz?

\ 
\ 

2.  Based on what you know about the stability of the cation-oxygen bonds in the BIG 8 (Lecture 1.5), do you think Feldspars are more or less stable (or resistant to weathering) than Quartz?

3.  In the phyllosilicates, how are the Si-tetrahedral sheets and Al-octahedral sheets linked?

4.  Why is kaolinite called a 1:1 clay mineral?

5.  Why are Micas called 2:1 clay minerals?

6.  What occupies the space between layers in Mica or Illite?

7.  Are there any other atoms besides Si in the tetrahedral sheets of a Mica or Illite? If so what element?

Note how water (WAT) and other cations (in this case Na) occupy the interlayer of the Vermiculites and Smectites. Unlike Kaolinite and Micas/Illites, Vermiculite/Smectites are shrinking and swelling (or expanding) clays because they have a weak enough negative charge that hydrated ions and water can enter the interlayer space. Smectites can expand even more than vermiculites because they have a lower negative charge so the positively charged cations that hold the layers together are few and the layers can disperse far apart.

## INVESTIGATION B: Clay Mineral Block Structures

Draw the mineral structure of the two clay minerals listed below using your lecture notes. Your drawing should include: tetrahedral sheets, octahedral sheets and labeled interlayers.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Clay <- c("Kaolinite (1:1)", "Smectite (2:1)")
Drawing <- c("","")

claydrawing <- data.frame(Clay, Drawing)
colnames(claydrawing) = c("Clay", "Drawing")

col.num = 2
row.num = 3 #including header

################
## SET AS HUXTABLE TYPE
################
claydrawing_ht <- as_hux(claydrawing)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = claydrawing_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

Explain why Kaolinite does not expand and Smectite does expand.

Draw a picture to aid your explanation.

## INVESTIGATION C: Clay Mineral Charges

1.  Insert the electrodes (which are connected to the battery) into the clay slurry.
2.  Wait \~ 1-2 minutes.
3.  Pull the electrodes out and determine the charge (positive or negative) of the clay mineral in the beaker. Wires are attached to the electrodes of the battery -- black is positive and white is negative. The clay will be attracted to the opposite charge of the battery electrode (positive attracted to negative).
4.  Clean the wires off once you are done with this investigation.

*Observations:*

What battery electrode positive (+) or negative (-) were the clays in the slurry attracted to?

What does this tell you about the charge on those clay particles?

## INVESTIGATION D: Clay Charge, Ion Charge, and Flocculation

1.  Shake up each of the 3 pre-made tubes containing a slurry of a clay-enriched B horizon in eastern MN with 20ml of either deionized H2O, NaCl, or AlCl3. The deionized H2O contains no cations, the NaCl solution contains Na+ ions, and the AlCl3 solution contains Al3+ ions.
2.  Wait 5 minutes and watch each tube for the formation of colloids (clay particles held together by electrical attraction to an ion) -- when they form, you should be able to see them with your naked eye. The clay colloids will start to settle to the bottom, whereas in the absence of colloid formation the clays will stay in suspension.

Record your observations in the table below. PAY ATTENTION TO THE CLARITY OF THE LIQUID AT THE TOP OF EACH TUBE OF SLURRYS AFTER 5 MINUTES. High clarity indicates the absence of suspended clay particles, while cloudy liquid indicates that clay particles are still floating around in suspension and have not settled out.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Slurry <- c("DI-H20", "NaCl", "AlCl3") 
Ion <- c("None", "Na+", "Al3+")
WaterClarity <-c("","","")

slurryion <- data.frame(Slurry, Ion, WaterClarity)

colnames(slurryion) = c("Solution", "Ion", "Water Clarity")

################
## SET AS HUXTABLE TYPE
################
slurryion_ht <- as_hux(slurryion)

################
## SET FORMATTING
################

tab1.1 = slurryion_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(2) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:3, col = 1:3, value = 'center') %>%
      #set_valign(row = 2:3, col = 1:2, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(2:3, 1:2, 60) %>%
      #set_top_padding(2:3, 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(kableExtra)

Slurry <- c("DI-H20", "NaCl", "AlCl3") 
Ion <- c("None", "Na+", "Al3+")
WaterClarity <-c("","","")

tab.1.1 <- data.frame(Slurry, Ion, WaterClarity)

colnames(tab.1.1) = c("Slurry", "Ion", "Water Clarity")

tab.1.1 %>% kbl(caption = "Observing Clay Flocculation", align = "ccc") %>% row_spec(0, bold = TRUE) %>% kable_classic(full_width = F, html_font = "Cambria")
```

**Interpret the results of Investigation D based on what you know about the charge on clays and the ionic potential of cations. What accounts for the differences observed?** Hint: Look at the lecture slides on clay mineralogy. The answer has to do with the charge on clay particles and the strength of the charge on the cation in solution.

## INVESTIGATION E: Shrink/Swell Observation

1.  Take two Dixie cups. Label one "Kaolinite" and the other "Smectite".
2.  Place a teaspoon of Wyoming bentonite in the "Smectite" cup (WY Bentonite is a type of smectite often used in engineering applications) and a teaspoon of Kaolinite in the "Kaolinite" cup.
3.  Using the graduated cylinder, add 20 ml of distilled H2O to each cup and stir with the pencil provided. (If the clay is not saturated, continue to add water in 10 ml increments and stir until you see excess water on the bottom).

Describe what happens to each of the clays. How much water (ml) could you add to each before there was excess water?

Explain this below using what you know about the properties of the major clay mineral groups and the difference between Kaolinites and Smectites.

## INVESTIGATION F: Clay Soils

Two soils have been set out for you to look at. One is dominated by Kaolinite clays and one by Smectite clays. Both soils have a clay percentage greater than 35%. Unlike the Kaolinite and Smectite clays in Investigation E, which are from mined geological deposits, these are real soil materials (dominated by each of these different clay types) and so they have color associated with them which is not due to the clay minerals alone.

Note the wetted and dried samples of each and describe what you see in terms of its behavior. Describe the characteristics of each.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(kableExtra)

DominantClayMineralinSoil = c("") 
MoistCharacteristics = c("")
DryCharacteristics = c("")
GeneralLocationsofSoilinUS = c("")

tab.1.1 <- data.frame(DominantClayMineralinSoil, MoistCharacteristics, DryCharacteristics, GeneralLocationsofSoilinUS) 
colnames(tab.1.1) = c("Dominant Clay Mineral in Soil", "Moist Characteristics", "Dry Characteristics", "General Locations of Soil in US)")

tab.1.1 %>% kbl(caption = "Examples of Clay Soils", align = "ccc") %>% row_spec(0, bold = TRUE) %>% kable_classic(full_width = F, html_font = "Cambria")
```

## INVESTIGATION G: Article

Read the short article provided. What was one of the clay minerals that solved the double murder mystery?
