# **Bulk Density, Porosity, and Organic Matter**

::: callout-note
## Objectives

-   Understand bulk density, particle density, and porosity and make appropriate calculations for each.
-   Understand the relationship between soil color and organic matter.
-   Understand the relationship between bulk density, texture, and organic matter.
:::

::: callout-tip
## Key Words & Concepts

-   Bulk density
-   Porosity
-   Particle density
-   WFPS
-   Sapric
-   Hemic
-   Fibric
:::

## INVESTIGATION A: Bulk Density Measurements

The soil core method is the one of the most commonly utilized methods for determining bulk density. A core is pounded into the soil and oven dried. We can then derive the bulk density from the oven dry mass of the soil and the known volume of the core. The graduated cylinder method is done in this lab to demonstrate another way to determine the bulk density of soil materials.

### Bulk Density: Soil Core Method

$$
D_{b,core}=\frac{(weight\ of\ core\ and\ soil\ -\ weight\ of\ core)\ g}{(volume\ of\ soil)\ cm^{3}}
$$

Volume of a cylinder (the soil core) = *(π × r^2^ × h)*. For π, use 3.14; for r (radius), measure the inside diameter of the core in cm and divide by 2; for h, measure the height of the core in cm.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Core weight (empty core provided)", "Weight of core and soil", "Weight of soil", "Core volume", "Bulk density of soil")
Mass <- c("grams", "grams", "grams", "cm3", "grams/cm3")

measurements <- data.frame(Measurement, Mass)
colnames(measurements) = c("", "Measurements")

col.num = 2
row.num = 6 #including header

################
## SET AS HUXTABLE TYPE
################
measurements_ht <- as_hux(measurements)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = measurements_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1, col = 1:2, value = 'center') %>%
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 2:row.num, col = 2, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

### Bulk Density: Graduated Cylinder Method

This method relies on filling a known volume with a soil material and is really only appropriate to use for loose (structureless single-grained) sands. Other soil types have aggregates and pore architectures that would make this method highly inaccurate.

::: callout-note
## Watch this video before you start this investigation and Investigation B - Direct Method.

```{r, echo = FALSE, message = FALSE, warning = FALSE, out.width = "25%"}
library(qrcode)

hydrometer_test_video <- qrcode::qr_code("https://youtu.be/od3UeU8XwtU")
#print(hydrometer_test_video)
plot(hydrometer_test_video)
```
:::

$$
D_{b,cyl}=\frac{(mass\ of\ cylinder\ and\ soil\ -\ mass\ of\ cylinder)\ g}{(volume\ of\ soil)\ cm^{3}}
$$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Weight of clean, dry 100 mL graduated cylinder", "Weight of clean, dry 100 mL graduated cylinder filled with 100 mL of fine sand", "Weight of soil", "Volume of soil (1 mL in cylinder = 1 cm3)", "Bulk density of soil")
Mass <- c("grams","grams", "grams", "cm3", "grams/cm3")

measurementsgradcyl <- data.frame(Measurement, Mass)
colnames(measurementsgradcyl) = c("", "Measurements")

col.num = 2
row.num = 6 #including header

################
## SET AS HUXTABLE TYPE
################
measurementsgradcyl_ht <- as_hux(measurementsgradcyl)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = measurementsgradcyl_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1, col = 1, value = 'center') %>%
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 2:row.num, col = 2, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

**NOTE: Pour the sand into a small beaker and save for Investigation B, Method 1.**

## INVESTIGATION B: Porosity

Soil porosity is the amount of pore space in a soil. Pore spaces are found between soil particles and are filled with either air or water. Soil porosity can be measured directly and indirectly.

### Soil Porosity: Direct Method

This method measures the amount of water it takes to fill the soil pores (percent porosity). This method only works for sands where the pores are large so we can fill them up just by pouring water in. For finer-textured soils this does not work very well because air gets trapped in very small pores or cannot penetrate into all of the spaces in the aggregates.

$$
Porosity(P)_{direct}=\frac{(volume\ of\ all\ pores)\ cm^{3}}{(total\ volume\ of\ soil)\ cm^{3}}\ *\ 100%
$$

1.  Fill a 100 mL graduated cylinder to the 100 mL mark with tap water. Tilt the beaker with the fine sandy soil (from Investigation A, Method 2) to a 15˚ angle. Pour the water very slowly into the soil-filled beaker until the water is **just at** the surface of the soil. The soil surface should glisten. Be very careful not to stir or shake the soil in the beaker.
2.  Let the soil stand for several minutes until all of the water soaks into the very small pores. After the water soaks in, add a little more water. Repeat this step as necessary.
3.  Determine the amount of water used to fill the pore space in the 100 cm^3^ of soil by subtracting the volume of water in the graduated cylinder from the original 100 mL volume. Note: We are assuming that the water poured into the soil exactly filled the pore space and thus equals the total pore volume.
4.  Dispose of wet soil in the bucket labeled "For Wet Soil Only."
5.  Record your results in the table below and make the necessary calculations.
6.  Rinse the beaker in the bucket provided and wipe it out with a towel.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Initial volume of water", "Volume of water remaining in cylinder after all soil pores are filled", "Volume of water filling soil pores (initial – remaining)", "Total volume of soil (1 cm3 = 1 mL)", "Porosity = (volume of water filling pores / total volume) × 100%")
Volume <- c("100 mL","mL", "mL", "100 mL", "%")

measurementsporositydirect <- data.frame(Measurement, Volume)
colnames(measurementsporositydirect) = c("", "Measurements")

col.num = 2
row.num = 6 #including header

################
## SET AS HUXTABLE TYPE
################
measurementsporositydirect_ht <- as_hux(measurementsporositydirect)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = measurementsporositydirect_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 1:2, col = 2, value = 'center') %>%
      set_align(row = 3:4, col = 2, value = 'right') %>%
      set_align(row = 5, col = 2, value = 'center') %>%
      set_align(row = 6, col = 2, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

{{< pagebreak >}}

### Soil Porosity: Indirect Method

This method uses bulk density and particle density to determine porosity (percent pore space). Most mineral soils have a particle density of about 2.6 to 2.7 grams/cm^3^. *For this exercise, assume a particle density of 2.65 g/cm^3^* (the same density as quartz, a major constituent of mineral soils).

$$
Porosity(P)_{indirect} =\ [{1 - (\frac{(D_{b}\ by\ graduated\ cylinder\ method)\ \frac{g}{cm^{3}}}{(2.65 \ \frac{g}{cm^{3}})})}]\ *\ 100%
$$

Depending on accuracy in collecting your porosity data, your calculations of porosity by the direct method and the indirect method should be similar. Record and compare your results.

Direct method: $$\\[0.1in]$$ Indirect method:

## INVESTIGATION C: Bulk Density of Mineral Soils

Soil bulk density is also highly correlated with soil texture and organic matter content. Each of the tupper- wares has been filled with the same volume of a range of materials. **Lift up each tupperware and do your best to estimate their relative masses, ordering them from lowest bulk density (lightest) to highest bulk density (heaviest).** Write down the box number and material (provided on box label) in the right column of the table below.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Rankings <- c("Lowest", "|", "|", "|", "|", "|", "Highest")
BoxNumber <- c("", "", "", "", "", "", "")

bulkdensityranked <- data.frame(Rankings, BoxNumber)
colnames(bulkdensityranked) = c("Rank", "Box Number")

col.num = 2
row.num = 8 #including header

################
## SET AS HUXTABLE TYPE
################
bulkdensityranked_ht <- as_hux(bulkdensityranked)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = bulkdensityranked_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 1:row.num, col= 1:2, 60) %>%
      #set_top_padding(row = 1:row.num, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

Give a general description of why the soil materials are ranked that way. Using your lecture notes, recall what factors influence bulk density.

$$\\[0.5in]$$

## INVESTIGATION D: Soil Organic Matter Estimation from Color

You can use soil color to roughly estimate the amount of organic matter within a soil. Use the Munsell color charts and the estimation cards to determine soil color and the subsequent amount of soil organic matter.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
SoilName <- c("A", "B", "C")
SoilColor <- c("", "", "")
OrganicMatter <- c("", "", "")

soilnamecolor <- data.frame(SoilName, SoilColor, OrganicMatter)
colnames(soilnamecolor) = c("Soil Name", "Soil Color", "Organic Matter (%)")

col.num = 3
row.num = 4 #including header

################
## SET AS HUXTABLE TYPE
################
soilnamecolor_ht <- as_hux(soilnamecolor)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = soilnamecolor_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      set_bottom_padding(row = 2:4, col= 1:3, 10) %>%
      #set_top_padding(row = 2:4, col = 1:3, 40) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

## INVESTIGATION E: Organic Soil Types and Experiential Observation

### Differences in Decomposition State of Organic Material

Examine the three different examples of organic soils and classify the three unknowns as sapric, hemic, or fibric.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Dish <- c("1", "2", "3")
Classification <- c("","", "")

dishclassification <- data.frame(Dish, Classification)
colnames(dishclassification) = c("Dish", "Classification")

col.num = 2
row.num = 4 #including header

################
## SET AS HUXTABLE TYPE
################
dishclassification_ht <- as_hux(dishclassification)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = dishclassification_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      #set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:4, col= 1:2, 40) %>%
      #set_top_padding(row = 2:4, col = 1:2, 40) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

### Experiential Observation of Organic Soil Material

Pick up a chunk of the moist organic soil in the bucket. This soil is a decomposed peat from northern Minnesota. Rub the material between your fingers and notice how most of it rubs away or disappears. In a mineral soil, you would feel the mineral grains (sand, silt or clay).

What is the definition of soil texture?

$$\\[0.3in]$$

Does this soil fit into any of our textural classes?
