---
editor: 
  markdown: 
    wrap: 72
---

# **Bulk Density, Porosity, and Organic Matter**

::: callout-note
## Objectives

-   Understand bulk density, particle density, and porosity and make
    appropriate calculations for each.
-   Understand the relationship between soil color and organic matter.
-   Understand the relationship between bulk density, texture, and
    organic matter.
:::

::: callout-tip
## Key Words & Concepts

-   Bulk density
-   Porosity
-   Particle density
-   WFPS
-   Sapric
-   Hemic
-   Fibric
:::

## INVESTIGATION A: Bulk Density Measurements

Bulk density can be determined in three ways: the clod method, the
graduated cylinder method, and the soil core method. The **soil core
method** is the most commonly utilized method and the one you should be
the most familiar with, given appropriate information on a test. The
clod method and graduated cylinder method are done in this lab purely to
demonstrate other ways to determine the bulk density of soil materials.

### 1. Bulk Density: Clod Method

$$
D_{b,clod}=\frac{(weight\ of\ clod\ in\ air)\ g}{(weight\ of\ clod\ in\ air\ -\ weight\ of\ clod\ in\ water)\ g*1cm^{3}/1g}
$$ Bulk density is defined as mass (oven dry weight) divided by total
volume. Measuring the mass of the clod is straightforward; measuring the
volume is not. We can use Archimedes' Principle to measure the volume of
any solid, regardless of its shape. Simply stated, if a 50g clod weighs
20g when immersed in water, it must have displaced 30g of water. Since 1
cm3 of water weighs 1 gram, the 30g of water is equivalent to 30 cm3 of
water. The volume of the clod, therefore, is 30 cm3. A general
mathematical expression for the above statement is:

Volume of clod (in cm3) = \[(weight of clod in air) -- (weight of clod
in water)\] g × 1cm3/1 g

This formula represents the denominator in the Db,clod formula above,
giving the result of your calculation in units of g/cm3 (also written as
g cm-3), which is bulk density.

**WATCH VIDEO TUTORIAL ON IPADS!!**

1.  Take one clod from the counter, a small tag will be attached
    indicating the clod ID (take note of this when filling out the chart
    on page 2).
2.  Measure the weight of the clod **in air** by attaching the paperclip
    to the hook hanging from the bottom of the scale. Slide the weights
    along the scale from left to right, starting with the largest
    weights first. Ask a TA for help if you need any assistance.
3.  Place the water bucket beneath the clod, allowing it to be submerged
    in the water while hanging from the paper clip and take the weight a
    second time. This will be the weight of the clod when **submerged in
    water**.
4.  Determine the volume (recall the formula provided above).
5.  Place the clod back on the counter, and set the scale back to 0 for
    the next student.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Weight of clod in air", "Weight of clod submerged in water", "Volume of water displaced by clod", "Volume of clod", "Bulk density")
ClodA <- c("grams","grams", "cm3", "cm3", "grams/cm3")
ClodB <- c("grams","grams", "cm3", "cm3", "grams/cm3")

clodclod <- data.frame(Measurement, ClodA, ClodB)
colnames(clodclod) = c("", "Clod __", "Clod __")

col.num = 3
row.num = 6 #including header

################
## SET AS HUXTABLE TYPE
################
clodclod_ht <- as_hux(clodclod)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hashtag in front of them

tab1.1 = clodclod_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1, col = 1:2, value = 'center') %>%
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 2:row.num, col = 2:3, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      set_bottom_padding(row = 1, col= 1:3, 20) %>% #trying to add roughly one line of space below the header row to match the existing manual
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

### 2. Bulk Density: Soil Core Method

$$
D_{b,core}=\frac{(weight\ of\ core\ and\ soil\ -\ weight\ of\ core)\ g}{(volume\ of\ soil)\ cm^{3}}
$$

*Volume of a cylinder (the soil core) = (π × r^2^ × h)* For r (radius),
measure the *inside* diameter of the core in centimeters (**not inches**).

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Soil core #", "Core weight (empty core provided)", "Weight of core and soil", "Weight of soil", "Core volume", "Bulk density of soil")
Mass <- c("","grams", "grams", "grams", "cm3", "grams/cm3")

measurements <- data.frame(Measurement, Mass)
colnames(measurements) = c("", "Measurements")

col.num = 2
row.num = 7 #including header

################
## SET AS HUXTABLE TYPE
################
measurements_ht <- as_hux(measurements)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = measurements_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1, col = 1:2, value = 'center') %>%
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 2:row.num, col = 2, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

### 3. Bulk Density: Graduated Cylinder Method

$$
D_{b,cyl}=\frac{(mass\ of\ cylinder\ and\ soil\ -\ mass\ of\ cylinder)\ g}{(volume\ of\ soil)\ cm^{3}}
$$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Weight of clean, dry 100 mL graduated cylinder", "Weight of clean, dry 100 ml graduated cylinder filled with 100 ml of fine sand", "Weight of soil", "Volume of soil (1 ml in cylinder = 1 cm3)", "Bulk density of soil")
Mass <- c("grams","grams", "grams", "cm3", "grams/cm3")

measurementsgradcyl <- data.frame(Measurement, Mass)
colnames(measurementsgradcyl) = c("", "Measurements")

col.num = 2
row.num = 6 #including header

################
## SET AS HUXTABLE TYPE
################
measurementsgradcyl_ht <- as_hux(measurementsgradcyl)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = measurementsgradcyl_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1, col = 1, value = 'center') %>%
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 2:row.num, col = 2, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

NOTE: Pour the sand into a small beaker and save for Investigation B,
Method 1.

## INVESTIGATION B: Porosity

Soil porosity is the amount of pore space in a soil. Pore spaces are
found between soil particles and are filled with either air or water.
Soil porosity can be measured directly and indirectly.

### 1. Soil Porosity: Direct Method

This method measures the amount of water it takes to fill the soil pores
(percent porosity). This method only works for sands where the pores are
large so we can fill them up just by pouring water in. For
finer-textured soils this does not work very well because air gets
trapped in very small pores.

$$
Porosity(P)_{direct}=\frac{(volume\ of\ all\ pores)\ cm^{3}}{(total\ volume\ of\ soil)\ cm^{3}}\ *\ 100%
$$

1.  Fill a 100 ml graduated cylinder to the 100 ml mark with tap water.
    Tilt the beaker with the fine sandy soil (from Investigation A,
    Method 2) to a 15˚ angle. Pour the water very slowly into the
    soil-filled beaker until the water is **just at** the surface of the
    soil. The soil surface should glisten. Be very careful not to stir
    or shake the soil in the beaker.
2.  Let the soil stand for several minutes until all of the water soaks
    into the very small pores. After the water soaks in, add a little
    more water. Repeat this step as necessary.
3.  Determine the amount of water used to fill the pore space in the 100
    cm3 of soil by subtracting the volume of water in the graduated
    cylinder from the original 100 ml volume. Note: We are assuming that
    the water poured into the soil exactly filled the pore space and
    thus equals the total pore volume.
4.  Dispose of wet soil in the bucket labeled "For Wet Soil Only"
5.  Record your results in the table below and make the necessary
    calculations.
6.  Rinse the beaker in the bucket provided and wipe it out with a
    towel.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Measurement <- c("Initial volume of water", "Volume of water remaining in cylinder after all soil pores are filled", "Volume of water filling soil pores (initial – remaining)", "Total volume of soil (1 cm3 = 1 ml)", "Porosity = (volume of water filling pores / total volume) × 100%")
Volume <- c("100 mL","mL", "mL", "100 mL", "%")

measurementsporositydirect <- data.frame(Measurement, Volume)
colnames(measurementsporositydirect) = c("", "Measurements")

col.num = 2
row.num = 6 #including header

################
## SET AS HUXTABLE TYPE
################
measurementsporositydirect_ht <- as_hux(measurementsporositydirect)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = measurementsporositydirect_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1, value = 'center') %>%
      set_align(row = 1:2, col = 2, value = 'center') %>%
      set_align(row = 3:4, col = 2, value = 'right') %>%
      set_align(row = 5, col = 2, value = 'center') %>%
      set_align(row = 6, col = 2, value = 'right') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:3, col= 1:2, 60) %>%
      #set_top_padding(row = 2:3, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

### 2. Soil Porosity: Indirect Method

This method uses bulk density and particle density to determine porosity
(percent pore space). Most mineral soils have a particle density of
about 2.6 to 2.7 grams/cm3. *For this exercise, assume a particle
density of 2.65 g/cm3* (the same density as quartz, a major constituent
of mineral soils).

$$
Porosity(P)_{indirect} =\ ({1 - (\frac{(D_{b}\ by\ graduated\ cylinder\ method)\ \frac{g}{cm^{3}}}{(particle\ density)\ \frac{g}{cm^{3}}})})\ *\ 100%
$$

Depending on accuracy in collecting your porosity data, your
calculations of porosity by the direct method and the indirect method
should be similar. Record your results.

Direct method:

 

Indirect method:

## INVESTIGATION C: Bulk Density of Mineral Soils

Soil bulk density is the mass of oven dry soil (grams) divided by the
total volume of soil (cm3). Soil bulk density may be highly variable
over small distances and may even change within the same horizon. Bulk
density is also highly correlated with soil texture. Compare and rank
the bulk densities for the soil materials within each box from low to
high bulk density:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Rankings <- c("Lowest", "|", "|", "|", "|", "|", "Highest")
BoxNumber <- c("", "", "", "", "", "", "")

bulkdensityranked <- data.frame(Rankings, BoxNumber)
colnames(bulkdensityranked) = c("Rank", "Box Number")

col.num = 2
row.num = 8 #including header

################
## SET AS HUXTABLE TYPE
################
bulkdensityranked_ht <- as_hux(bulkdensityranked)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = bulkdensityranked_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 1:row.num, col= 1:2, 60) %>%
      #set_top_padding(row = 1:row.num, col = 1:2, 60) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

Give a general description of why the soil materials are ranked that
way. Using your lecture notes, recall what factors influence bulk
density.

## INVESTIGATION D: Soil Color and Soil Organic Matter

You can use soil color to estimate the amount of organic matter within a
soil. Use the Munsell color charts and the estimation cards to determine
soil color and the subsequent amount of soil organic matter.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
SoilName <- c("A", "B", "C")
SoilColor <- c("", "", "")
OrganicMatter <- c("", "", "")

soilnamecolor <- data.frame(SoilName, SoilColor, OrganicMatter)
colnames(soilnamecolor) = c("Soil Name", "Soil Color", "Organic Matter (%)")

col.num = 3
row.num = 4 #including header

################
## SET AS HUXTABLE TYPE
################
soilnamecolor_ht <- as_hux(soilnamecolor)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = soilnamecolor_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:4, col= 1:3, 40) %>%
      #set_top_padding(row = 2:4, col = 1:3, 40) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

## INVESTIGATION E: Organic Soil Types and Experiential Observation

Examine the three different examples organic soils and classify the
three unknowns as sapric, hemic, or fibric.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(huxtable)
library(dplyr)

################
## BUILD TABLE
################
Dish <- c("1", "2", "3")
Classification <- c("","", "")

dishclassification <- data.frame(Dish, Classification)
colnames(dishclassification) = c("Dish", "Classification")

col.num = 2
row.num = 4 #including header

################
## SET AS HUXTABLE TYPE
################
dishclassification_ht <- as_hux(dishclassification)

################
## SET FORMATTING
################
#NOTE: only thing you really need to mess with down here is the padding. If you need extra row height, keep them - if not, just put a hastag in front of them

tab1.1 = dishclassification_ht %>% 
      set_all_padding(4) %>% 
      set_outer_padding(0) %>% 
      set_number_format(0) %>% #this tells it how many decimals to print after any numbers
      set_bold(row = 1, col = everywhere) %>% #make header row bold
      #set_bottom_border(row = 1, col = everywhere) %>% # probably dont need this because you set all borders below
      set_width(0.9) %>% #; 1 == full width
      set_align(row = 1:row.num, col = 1:col.num, value = 'center') %>%
      set_valign(row = 1:row.num, col = 1:col.num, value = 'middle') %>%
      #set_number_format(2:3, 1, value = NA) %>% #; don't format these cells as numbers
      #set_col_width(c(0.3, 0.6)) %>%
      set_latex_float(value = "h!") %>%
      #set_row_height(1:3, c(0.02, 0.2, 0.2)) %>%
      #set_bottom_padding(row = 2:4, col= 1:2, 40) %>%
      #set_top_padding(row = 2:4, col = 1:2, 40) %>%
      #set_row_height(c(0.02, 0.2, 0.2)) %>%
      set_all_borders(brdr(1, "solid", "black"))
      #set_caption("Pots of jam for sale") # this is if we want a caption - I don't think we need any captions but leaving it here just in case

tab1.1
```

Pick up a chunk of the moist organic soil in the bucket. This soil is a
decomposed peat from northern Minnesota. Rub the material between your
fingers and notice how most of it rubs away or disappears. In a mineral
soil, you would feel the mineral grains (sand, silt or clay).

What is the definition of soil texture?

   

Does this soil fit into any of our textural classes?

   

## INVESTIGATION F: Ranges and Estimation of Bulk Soil Physical Properties

*There is no lab station*; find a spot near the lab and answer the
following questions. Use your lecture notes (in particular, **Lecture
1.4**)!

1.  What is the common Db range (g/cm3) for typical cultivated topsoils?

   

2.  What is the Db range (g/cm3) for typical uncultivated topsoils?

   

3.  I have two soils. Soil 1 is a clay with a Db of 1.4 g/cm3. Soil 2 is
    a loamy sand with a Db of 1.4 g/cm3. Will root penetration likely be
    inhibited in either of these soils? Why or why not?

   

4.  Someone tells you they calculated a Db of 2.7 g/cm3 from a soil core
    they took. Should you believe them? Why or why not?

   

5.  In general, what is a reasonable value to guess for WFPS (i.e., if
    you don't know anything else about a soil)?

   

6.  Soil A has a porosity of 52% and a WFPS of 92%. Soil B has a
    porosity of 23% and a WFPS of 92%. Which one contains more water?

   

7.  Without doing any calculations, which soil has the higher bulk
    density? Why?

   
